Score-based tests of measurement invariance: Use in practice

### 2014-04-23 frontiers/frontiers.Rnw
Final revisions to frontiers paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@334 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Many of the papers in this special issue focus on the topic of approximate
measurement invariance: we know that strict hypotheses of measurement invariance
do not hold exactly across different groups, and this should be reflected in
corresponding tests of measurement invariance.  Under a Bayesian
approach, we may utilize the idea of approximate invariance
\cite<e.g.,>{muthen13}, whereby across-group equality constraints on
parameters are replaced with informative prior distributions.  In this paper, we
describe an alternative approach: the development of test statistics that 
are especially sensitive to violations that are monotonic w.r.t. the
variable of interest (and 
less sensitive to violations that are non-monotonic w.r.t. the
variable of interest).  Because monotonic violations are more likely to be
interpretable and interesting to the researcher, we can gain more
power to detect these violations by de-emphasizing other types of violations.
The resulting test statistics are specifically applicable to
situations where one  
wishes to test for measurement invariance with respect to an ordinal variable, 
and they are special cases of a family of tests that may be used to study 
measurement invariance w.r.t.\ continuous, categorical, and ordinal
variables.

The study of measurement invariance w.r.t.\ categorical auxiliary 
variables (via, e.g., likelihood ratio tests) is popular and well
known, and ordinal auxiliary variables are typically treated as
categorical in measurement invariance contexts.
The study of measurement invariance w.r.t.\ continuous
variables is newer and less known: along with the family
described here, other methods include moderated factor
models \cite{bauhus09,moldol10,pur02} and factor mixture models
\cite{dolmaa98,lubmut05}.  These methods require estimation of a model
of greater complexity, while the tests described in this paper work solely
on the output of a traditional factor model \cite<see>[for
further comparison of these methods]{MerZei13}.

The family of tests described here have recently been applied
to the study of measurement invariance \cite{MerZei13,MerFanZei}, though their
practical application has been limited to a small set of simulations and
data examples.  In this paper, we provide a detailed illustration of
the tests' use and performance under
scenarios likely to be encountered in practice.  While the previous
papers have described and studied the tests under ideal conditions, we
focus here on two topics of interest to the applied researcher:
software considerations for carrying out the tests, and test
performance under model misspecification.
%In order to fix the psychometric scale, we need to specify the violating parameter.  
%Moreover, in practice model misspecification is an inevitable issue to a certain extent 
%and may interfere with parameter invariance specificity. A family of tests derived from 
%stochastic process has been recently proposed to identify specific parameters impacted 
%by measurement invariance violation. However, these tests have only been used to a limited 
%type of parameter violation. It is this paper's intent to provide a detailed examination 
%of the tests' performance under practical research scenarios. 
In the following sections, we first briefly review the  theoretical framework of
the proposed tests and provide a short tutorial illustrating the use of the
tests in \proglang{R} \cite{R11}. Next, we study the tests' performance in simulations that mimic
practical research scenarios.  Finally, we provide some further discussion on
the tests' use in practice.
% We acknowledge problems associated with $p$-values (e.g., corrections for multiple
% tests, researcher degrees of freedom, interpretations, etc) but
% ignore them throughout the paper for expository purposes.

@@

We focus on measurement invariance of the factor loadings associated with 
the GQ-6 scale, using a one-factor model.  This scale consists of five
items scored from 1--9.  While the ``age group'' variable against 
which we are testing measurement invariance is best considered ordinal, 
for demonstration we consider its treatment as categorical, continuous, and ordinal.  
Each of these treatments is described below in a separate section.

>>>>

Some of the papers in this special issue focus on the topic of approximate
measurement invariance: we know that strict hypotheses of measurement invariance
do not hold exactly across different groups, and this should be reflected in
corresponding tests of measurement invariance.  Under a Bayesian
approach, we may implement the idea of approximate invariance
\cite<e.g.,>{muthen13} by replacing across-group equality constraints on
parameters with informative prior distributions.  In this paper, we
describe an alternative approach: the development of test statistics that 
are especially sensitive to violations that are monotonic w.r.t. the
variable of interest (and 
less sensitive to violations that are non-monotonic w.r.t. the
variable of interest).  Because monotonic violations are more likely to be
interpretable and interesting to the researcher, we can gain more
power to detect these violations by de-emphasizing other types of violations.
The resulting test statistics are specifically applicable to
situations where one  
wishes to test for measurement invariance with respect to an ordinal variable, 
and they are special cases of a family of tests that may be used to study 
measurement invariance w.r.t.\ continuous, categorical, and ordinal
variables.

The study of measurement invariance w.r.t.\ categorical auxiliary 
variables (via, e.g., likelihood ratio tests) is popular and well
known, and ordinal auxiliary variables are typically treated as
categorical in measurement invariance contexts.
The study of measurement invariance w.r.t.\ continuous
variables is newer and less known: along with the family
described here, other methods include moderated factor
models \cite{bauhus09,moldol10,pur02} and factor mixture models
\cite{dolmaa98,lubmut05}.  These methods require estimation of a model
of greater complexity, while the tests described in this paper work solely
on the output of a traditional factor model \cite<see>[for
further comparison of these methods]{MerZei13}.  These methods all
assume that the estimated model is correctly specified, save
possibly for differences in parameter values between individuals.

The family of tests described here have recently been applied
to the study of measurement invariance \cite{MerZei13,MerFanZei}, though their
practical application has been limited to a small set of simulations and
data examples.  In this paper, we provide a detailed illustration of
the tests' use and performance under
scenarios likely to be encountered in practice.  While the previous
papers have described and studied the tests under ideal conditions, we
focus here on two topics of interest to the applied researcher:
software considerations for carrying out the tests, and test
performance under model misspecification.  The latter issue is
particularly important because, in practice, all models are
misspecified.  Hence, practically-useful tests of measurement
invariance should be robust to model misspecification.
%In order to fix the psychometric scale, we need to specify the violating parameter.  
%Moreover, in practice model misspecification is an inevitable issue to a certain extent 
%and may interfere with parameter invariance specificity. A family of tests derived from 
%stochastic process has been recently proposed to identify specific parameters impacted 
%by measurement invariance violation. However, these tests have only been used to a limited 
%type of parameter violation. It is this paper's intent to provide a detailed examination 
%of the tests' performance under practical research scenarios.

In the following sections, we first briefly review the  theoretical framework of
the proposed tests and provide a short tutorial illustrating the use of the
tests in \proglang{R} \cite{R11}. Next, we study the tests' performance in simulations that mimic
practical research scenarios.  Finally, we provide some further discussion on
the tests' use in practice.
% We acknowledge problems associated with $p$-values (e.g., corrections for multiple
% tests, researcher degrees of freedom, interpretations, etc) but
% ignore them throughout the paper for expository purposes.

@@

We focus on measurement invariance of the factor loadings associated with 
one of the scales in the dataset, the GQ-6 scale \cite{mccemm02}.  This
scale consists of five Likert scale items (there is a sixth item that
is omitted from analyses, following Froh et al.) with seven points
each.  We fit a one-factor model to these items, examining whether the
factor model parameters are invariant with respect to age group.  
While the age group variable is best considered ordinal, 
for demonstration we consider its treatment as categorical, continuous, and ordinal.  
Each of these treatments is described below in a separate section.
